From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting

### 2024-05-15 paper/paper.qmd
Polished theory part

We assume that the runoff for a specific point in time of all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ that are known for the past $\tau=1,\ldots N_\tau$ time instances, i.e.

@@

The fields are evaluated on a spatial domain $x=1,\ldots N_x$ and $y=1,\ldots N_y$ that is sufficiently large to capture all significant contributions to the modelled runoff.

We propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network as it is visualized in @fig-ConvLSTM_FC and described in detail in the following. In order to provide an overview, we will go through the main ingredients of this architecture one-by-one.

![](./ConvLSTM_FC.png){#fig-ConvLSTM_FC fig-align="center"}

@@

Before turning to the convolutional LSTM, the simpler architecture of the plain LSTM is examined and should serve as a role model for the later consideration of the full ConvLSTM.

The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically tailored for modeling temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$. The sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ and the endpoint should coincide with the specific element in the time series connected to time $t$, i.e. $\vec{X}^t[N_\tau] \equiv \vec{X}^t$. The $N_k$ is the number of used input "channels" and can, for example, represent different measurable quantities. Its unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).

![Inner structure of a Long Short-Term Memory Cell](./LSTM.png){#fig-lstm fig-align="center"}

This performance in modeling long-range dependencies has been validated in various studies. The key component of the LSTM's innovation is its memory cell, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as long-term memory. This state information complements the the so-called hidden state $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$ vector, which is also known from simpler neural network architectures. In case of the LSTM, the hidden state vector plays the role of the short-term meory. The cell state and the hidden state are vectors, where each element is associated with one of the so-called $N_h$ hidden layers, labeled by $h$. These two state vectors are determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$.

In particular, the forget gate $\vec{F}^t[\tau]$ defines the portion of the previous (long-term memory) cell state $\vec{C}^t[\tau-1]$ that should be kept​. The input gate $\vec{I}^t[\tau]$ controls the contribution of the current input that is used to update the long-term memory,$\vec{C}^t[\tau]$. The output gate $\vec{O}^t[\tau]$ then determines how much of this updated long-term memory contributes to the new (short-term memory) hidden state, $\vec{H}^t[\tau]$.

@@

For the sake of brevity, we write the mapping more compactly in matrix-vector form as $$
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g}\vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g}  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
$$ {#eq-LSTM-gates}

Second, the actual gate vectors are computed by the core equations of the LSTM as proposed by [@XXX]:

@@

The full action of the LSTM network up to the end of the sequence can therefore be written as a nested function call

@@

where $L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)$ represents @eq-LSTM-gates and @eq-LSTM and the initial conditions are chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$.

The final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ encode information on the full input sequence ending at time $t$. This information has to be decoded to obtain useful information via an appropriate subsequent layer, as it is described in the @sec-FC_layer.

@@

Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of, e.g., two-dimensional maps. To adress this limitation we employ a ConvLSTM architecture as described in the following.

In this kind of network the elements of the input sequence are given as spatially varying fields $\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}$, where $x\in[1, N_x]$ and $y \in [1, N_y]$ run over the horizontal and vertical dimensions of the map, respectively. In order to enable the "learning" of spatial patterns, the free parameters of the network are replaced by two-dimenmsional convolutional kernels $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}$ and $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}$. The width and the height of the kernels are given by $N_\xi$ and $N_\eta$, respectively and $\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]$, where we assume odd numbers for the kernel sizes.

@@

It becomes immediately apparent that in case of the ConvLSTM, gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well. We can write this mapping in the same fashion as @eq-LSTM-gates but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
$$
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g} \ast \vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g} \ast  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
$$

The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in @eq-LSTM but with all appearing quantities meaning standing now for vector fields instead of simple vectors.

@@

As stated in Sec. @sec-LSTM, the final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ of the ConvLSTM encode information on the full input sequence. In order to contract this information to the runoff vector $\vec{R}^t$ representing the $N_r$ rivers, we propose to subject the final short-term memory (i.e. the hidden state $\vec{H}^t[N_\tau]$) to an additional FC network.

In particular, the dimensionality of the vector field $\vec{H}^t[N_\tau]$ is sequentially reduced by three nested FC layers, each connected to the other by the Rectified Linear Unit (ReLU), see @fig-ConvLSTM_FC. Integrating over artificial degrees of freedom in a step-wise fashion has turned out to be benificial (@XXX).

Spelled out in mathematics the runoff of the $r$-th river is then obtained via (using Einstein's convention)

@@

where $a=1,\ldots N_a$ and $b=1,\ldots N_b$ and $N_h\cdot N_x\cdot N_y > N_a > N_b > N_r$ in order to achieve the aforementioned step-by-step reduction of dimensionality. The weights $\mathcal{W}$ and biases $\mathcal{B}$ stand for parameters that are optimized during the training of the network.

In matrix-vector notation this can compactified to $$
\begin{aligned}
\vec{R}^t = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{H}^t[N_\tau] + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ .
\end{aligned}
$$

Combining this equation with @eq-L_nested provides finally an explicit formula for the initial assumption of modelling the runoff for time $t$ as a functional of a sequence of atmospheric fields, i.e.

$$
\begin{aligned}
\vec{R}^t & = \vec{M}(\{X^t_k[x,y,\tau]\}) \\
& = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{L}_H \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (0, 0) \right) \ldots \right) \right) + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ ,
\end{aligned}
$$ where the $\vec{L}_H$ means that only the hidden state vector of the final ConvLSTM call is forwarded to the FC layer. In oder to make this model accurate the hyperparameters have to be chosen reasonable and the free parameters are to be adjusted accordingly.

## Technical details

>>>>

We assume that the runoff for a specific point in time for all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ that are known for the past $\tau=1,\ldots N_\tau$ time instances, i.e.

@@

The fields are evaluated on a spatial domain $x=1,\ldots N_x$ and $y=1,\ldots N_y$ that is sufficiently large to capture all significant local and non-local contributions of atmospheric fields to the river runoff.

Usually such a mapping is realized in a suitable hydrological model that simulates all relevant physical processes that transform variables like, e.g., precipitation and evaproration to the river runoff, relying on expertise for tuning all appearing parameters to reasonable values.

As an alternative, we propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network. That way, there is no need to have knowledge of involved processes and how they can be modeled, instead, these features can be "lerned" by the network in an automated manner. Our poroposed network architecture is visualized in @fig-ConvLSTM_FC and described in detail in the following. In order to provide an overview, we will go through the main ingredients of this architecture one-by-one.

![**Combined ConvLSTM and FC network architecture**. The starting point is the continuous timeseries of input data $\{\vec{X}^t\}$, upper white block. From this series, a contiguous sequence of $N_\tau$ elements (yellow block) is used to feed a chain of $N_\tau$ connected ConvLSTM cells (light blue block) building the ConvLSTM network (grey block). The input seuqence is mapped via weighting matrices $\pmb{\mathcal{M}}^g$ (green blocks) onto gate vectors $\vec{g}^t[\tau]$. The gate vectors are then used to update the cell state $\vec{C}^t[\tau-1]$ and the hidden state $\vec{H}^t[\tau-1]$ of the last ConvLSTM cell to the current values $\vec{C}^t[\tau]$ and $\vec{H}^t[\tau]$, respectively. The update is performed with LSTM core equations collectively described by the mapping L, see @eq-LSTM. The weighting matrices $\pmb{\mathcal{N}}^g$ (green blocks) control how much of the last hidden state enters the updated state. The final output of the ConvLSTM $\vec{H}^t[\tau]$ is then propagated to a FC network, which itself is a chain of three FC layers consisiting of weigthing matrices $\pmb{\mathcal{W}}$ and connected via ReLU functions, see @sec-FC_layer. The final result is then the river runoff $\vec{R}^t$ for all rivers considered at the current time instance $t$ (white block on the lower left). Note that all bias vectors are ommitted for the sake of clarity. See text for more information.](./ConvLSTM_FC.png){#fig-ConvLSTM_FC fig-align="center"}

@@

Before turning directly to the convolutional LSTM, the simpler architecture of the plain LSTM is examined and should serve as a role model for the later consideration of the full ConvLSTM.

The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically tailored for modeling temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$. The sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ and the endpoint should coincide with the specific element in the time series connected to time $t$, i.e. $\vec{X}^t[N_\tau] \equiv \vec{X}^t$, see @fig-ConvLSTM_FC. The $N_k$ is the number of used input "channels" and can, for example, represent different measurable quantities. Its unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).

![**Inner structure of a Long Short-Term Memory Cell**. See @fig-ConvLSTM_FC and text for information.](./LSTM.png){#fig-lstm fig-align="center"}

This performance in modeling long-range dependencies has been validated in various studies [@XXX]. The key component of the LSTM's innovation is its cell state, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as *long-term memory*. This state information complements the so-called hidden state $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$ vector, which is also known from simpler neural network architectures. In case of the LSTM, the hidden state vector plays the role of the *short-term meory*.

The cell state and the hidden state are vectors, where each element is associated with one of the $N_h$ hidden layers, labeled by $h$, which are internal, *artificial* degrees of freedom that enable the high adaptibility of neural networks. These two state vectors are determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$, see @fig-lstm for a visualization.

In particular, the forget gate $\vec{F}^t[\tau]$ defines the portion of the previous (long-term memory) cell state $\vec{C}^t[\tau-1]$ that should be kept, see dashed cyan box therin​. The input gate $\vec{I}^t[\tau]$ controls the contribution of the current input used to update the long-term memory, $\vec{C}^t[\tau]$ (magenta and yellow boxes). The output gate, $\vec{O}^t[\tau]$, then determines how much of this updated long-term memory contributes to the new (short-term memory) hidden state, $\vec{H}^t[\tau]$ (black dashed box).

@@

For the sake of brevity, we can write the mapping more compactly in matrix-vector form as $$
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g}\vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g}  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
$$ {#eq-LSTM-gates}

Second, the actual gate vectors are computed by the core equations of the LSTM as proposed by @hochreiter1997long:

@@

The full action of the LSTM network up to the end of the sequence can be written as a nested function call

@@

where $L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)$ represents @eq-LSTM-gates and @eq-LSTM. For the present work, the initial conditions are chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$, which simply means that there is no memory longer than $N_\tau$ time steps.

The final output of the ConvLSTM chain, $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$, encode information on the full input sequence ending at time $t$. This information has to be decoded to obtain useful information via an appropriate subsequent network, as it is described in the @sec-FC_layer.

@@

Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of, e.g., two-dimensional maps as atmospheric-ocean interface fields. To adress this limitation we employ a ConvLSTM architecture as described in the following.

In this kind of network the elements of the input sequence are given as spatially varying fields $\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}$, where $x\in[1, N_x]$ and $y \in [1, N_y]$ run over the horizontal and vertical dimensions of the map, respectively. In order to enable the "learning" of spatial patterns, the free parameters of the network are replaced by two-dimenmsional convolutional kernels $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}$ and $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}$. The width and the height of the kernels are given by $N_\xi$ and $N_\eta$, respectively and $\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]$, where, without loss of generality, we assume odd numbers for the kernel sizes.

@@

It becomes immediately apparent that in case of the ConvLSTM, the gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well. We can write this mapping in the same fashion as @eq-LSTM-gates but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
$$
\begin{aligned}
\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g} \ast \vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g} \ast  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
\end{aligned}
$$

The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in @eq-LSTM but with all appearing quantities now meaning vector fields instead of simple vectors.

@@

As stated in Sec. @sec-LSTM, the final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ of the ConvLSTM encode information on the full input sequence. In order to contract this information to obtain the runoff vector $\vec{R}^t$ representing the $N_r$ rivers, we propose to subject the final short-term memory (i.e. the hidden state $\vec{H}^t[N_\tau]$) to an additional FC network.

In particular, the dimensionality of the vector field $\vec{H}^t[N_\tau]$ is sequentially reduced by three nested FC layers, each connected to the other by the Rectified Linear Unit (ReLU), see @fig-ConvLSTM_FC. Integrating over artificial degrees of freedom in a step-wise fashion has turned out to be benificial \[@XXX\].

Spelled out in mathematics, the runoff of the $r$-th river is then obtained via (using Einstein's convention)

@@

where $a=1,\ldots N_a$, $b=1,\ldots N_b$ and the hyper parameters $N_a$ and $N_b$ are chosen such that $N_h\cdot N_x\cdot N_y > N_a > N_b > N_r$ in order to achieve the aforementioned step-by-step reduction of dimensionality. The weights $\mathcal{W}$ and biases $\mathcal{B}$ stand for parameters that are optimized during the training of the network.

In matrix-vector notation this can be compactified to $$
\begin{aligned}
\vec{R}^t = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{H}^t[N_\tau] + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ .
\end{aligned}
$$

Combining the last equation with @eq-L_nested provides finally an explicit formula for the initial assumption of modelling the runoff for time $t$ as a functional of a sequence of atmospheric fields, i.e.

$$
\begin{aligned}
\vec{R}^t & = \vec{M}(\{X^t_k[x,y,\tau]\}) \\
& = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{L}_H \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (0, 0) \right) \ldots \right) \right) + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ ,
\end{aligned}
$$ where the $\vec{L}_H$ means that only the hidden state vector of the final ConvLSTM call is forwarded to the FC layer.

In @sec-technical_details, we present the employed model and data setup as well as the choice for all mentioned hyper parameters that lead to an adequate model performance after training.

## Technical details {#sec-technical_details}
